---
title: "Genotype plus Genotype-By-Block-of-Environments Biplots"
author: "Kevin Wright"
date: "`r Sys.Date()`"
bibliography: gge.bib
output:
  html_document:
    highlight: tango
vignette: >
  % \VignetteIndexEntry{Genotype plus Genotype-By-Block-of-Environments Biplots}
  % \VignetteEncoding{UTF-8}
  % \VignetteEngine{knitr::rmarkdown}
---
<style type="text/css">
body, td { font-size: 14pt; color: #000; padding:1em; }
code, pre, #ws, #message { font-size: 12pt;
  color: #000; background-color: #f5f5f5; }
code {border: 1px solid #e0e0e0; padding: 0 5px;}
pre > code { border: 0; }
</style>

# Abstract

A short note about GGE and GGB biplots.

# R setup

```{r setup, message=FALSE}
library("knitr")
knitr::opts_chunk$set(fig.align="center", fig.width=6, fig.height=6)
options(width=90)
```

An example of a GGE (genotype plus genotype-by-environment) biplot similar to
figure 12 of @yan2006biplot.

```{r gge}
library(agridat)
data(yan.winterwheat)
dat1 <- yan.winterwheat

library(gge)
m1 <- gge(yield~gen*env, dat1, scale=FALSE)
biplot(m1, title="yan.winterwheat - GGE biplot",
       flip=c(1,0), origin=0, hull=TRUE)
```

@laffont2007numerical showed how to partition the sums-of-squares
simultaneously along the principal component axes and along 'G' and 'GxE'
axes.
```{r mosaic}
plot(m1, title="yan.winterwheat")
```

The mosaic plot above shows that the first principal component axis is
capturing almost all of the variation between genotypes, so that a projection
of the genotype markers onto the first principal component axis is a good
overall representation of the rankings of the genotypes.

@laffont2013genotype presented GGB (genotype plus genotype-by-block of
environments) biplots, which are useful to enhance the view of
mega-environments consisting of multiple locations.

```{r ggb}
library(agridat)
data(crossa.wheat)
dat2 <- crossa.wheat

# Define mega-environment groups of locations
dat2$eg <- ifelse(is.element(dat2$loc,
                             c("KN","NB","PA","BJ","IL","TC",
                               "JM","PI","AS","ID","SC","SS",
                               "SJ","MS","MG","MM")), "Grp1", "Grp2")

library(gge)
# Specify env.group as column in data frame
m2 <- gge(yield~gen*loc, dat2, env.group=eg, scale=FALSE)
biplot(m2, title="crossa.wheat - GGB biplot")
```

# References
